NumLin: Linear Types for Numerical Libraries

### 2018-10-21 write-up/paper/implementation.tex
Sketch out Implementation section

\section{Implementation}\label{sec:implementation}
\subsection{Implementation Strategy}

% TODO desugaring

@@

>>>>

\section{Implementation}\label{sec:implementation}

@@

\subsection{Implementation Strategy}

\lang\ transpiles to OCaml and its implementation follows the structure of a
typical domain-specific language compiler. Although \lang's current
implementation is not as embedded DSL, its the general design is simple enough
to adapt to being so and also to target other languages.

Alongside the transpiler, a `Read-Check-Translate' loop, benchmarking program
and a test suite are included in executable artifacts accompanying this paper.

\begin{enumerate}

    \item \textbf{Parsing}. A generated, LR(1) parser parses a text file into a
        syntax tree. In general, this part will vary for different languages
        and can also be dealt with using combinators or syntax-extensions (the
        EDSL approach) if the host language offers such support.

    \item \textbf{Desugaring}. The syntax tree is then desugared into a
        smaller, more concise, abstract syntax tree. This allows for the type
        checker to be simpler to specify and easier to implement.

    \item \textbf{Matrix Expressions} are also desugared into the abstract
        syntax tree through some simple pattern-matching.

    \item \textbf{Type checking}. The abstract syntax tree is explicitly typed,
        with some inference to make writing typical programs more convenient.

    \item \textbf{Code Generation}. The abstract syntax tree is translated into
        OCaml, with a few `optimisations' to produce more readable code. This
        process is type-preserving: \lang's type system is embedded into
        OCaml's (Figure~\ref{fig:type_grammar}), and so the OCaml type checker
        acts as a sanity check on the generated code.

\end{enumerate}

A very pleasant way to use \lang\ is to have the build system generate code at
\emph{compile-time} and then have it be used by other modules like normal OCaml
functions. This makes it possible and even easy to use \lang\ alongside
existing OCaml libraries; in fact, this is exactly how the benchmarking program
and test-suite use code written in \lang.

\begin{figure}[tp]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <fc>
            \alt `Z'
            \alt `S' <f>

            <t> ::= `'
            \alt `unit'
            \alt `bool'
            \alt `int'
            \alt `elt'
            \alt <f> `arr'
            \alt <f> `mat'
            \alt `!' <t>
            \alt \synt{$\forall$} <fc.> <t>
            \alt <t> \lit{$\otimes$} \synt{$t'$}
            \alt <t> \lit{$\multimap$} \synt{$t'$}
        \end{grammar}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{minted}[fontsize=\small]{ocaml}
module Arr =
  Owl.Dense.Ndarray.D

type z = Z
type 'a s = Succ

type 'a arr =
  A of Arr.arr
  [@@unboxed]

type 'a mat =
  M of Arr.arr
  [@@unboxed]

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}

\subsubsection{Desugaring}\label{subsubsec:desugaring}

Straightforward enough.

\subsubsection{Matrix Expressions}

Pattern matching!

\subsubsection{Type Checking and Fractional Permission Inference}

Unification!

\subsubsection{Code Generation}

A few examples?

### 2018-10-21 write-up/paper/lang_and_examples.tex
Sketch out Implementation section

The only argument is \highl{!x : !int}. The ! annotation on \texttt{x} is a
syntactic convenience for declaring the value to used intuitionistically, its
full and precise meaning is described in Section~\ref{sec:implementation}.

@@

With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
get very difficult to ensure memory is not being re-used incorrectly or re-used
enough whilst also enusring the aliasing assumptions for the BLAS/LAPACK
routine calls are guaranteed. However, using \lang\ makes incorrect aliasing
and memory re-use impossible. Additionally, \lang\ makes it very easy to spot
where memory is not being re-used enough \emph{precisely because} of its
explicit handling of matrix allocations, copies and frees.

>>>>

The only argument is \highl{!x : !int}. The ! annotation on \texttt{x} is a
syntactic convenience for declaring the value to used intuitionistically, its
full and precise meaning is described in Section~\ref{subsubsec:desugaring}.

@@

With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
get very difficult to ensure memory is not being re-used incorrectly/enough
whilst also enusring the aliasing assumptions for the BLAS/LAPACK routine calls
are guaranteed. However, using \lang\ makes incorrect aliasing and memory
re-use impossible. Additionally, \lang\ makes it possible to spot where memory
is not being re-used enough \emph{precisely because} of its explicit handling
of allocations, aliases and frees.
